---
title: "Mini Project #01: Fiscal Characteristics of Major US Public Transit Systems"
author: "Lavanya"
date: "2024-09-20"
execute:
  echo: false   # Hide the code in the final document
  eval: true    # Evaluate code chunks
  freeze: auto  # Ensure results are frozen when running
output:
  html_document:
    theme: flatly
---

Introduction
In this mini-project, I analyze the fiscal characteristics of major US public transit systems, focusing on farebox recovery, unlinked passenger trips (UPT), and vehicle revenue miles (VRM) for 2022. The data is sourced from the National Transit Database and includes various transit agencies and modes of transportation.

Methodology
The analysis is based on data from:

1) 2022 Fare Revenue report
2) Operating Expenses report
3) Monthly Ridership data for 2022

Data Cleaning and Merging

```{r}
# Load required libraries
library(tidyverse)
library(readxl)
library(lubridate)

# Fare revenue data
FARES <- read_xlsx("2022_fare_revenue.xlsx") %>%
  select(-`State/Parent NTD ID`, -`Reporter Type`, -`Reporting Module`, -`TOS`, 
         -`Passenger Paid Fares`, -`Organization Paid Fares`) %>%
  filter(`Expense Type` == "Funds Earned During Period") %>%
  select(-`Expense Type`) %>%
  group_by(`NTD ID`, `Agency Name`, `Mode`) %>%
  summarize(`Total Fares` = sum(`Total Fares`), .groups = 'drop')

# Operating expenses data
EXPENSES <- readr::read_csv("2022_expenses.csv") %>%
  select(`NTD ID`, `Agency`, `Total`, `Mode`) %>%
  mutate(`NTD ID` = as.integer(`NTD ID`)) %>%
  rename(Expenses = Total) %>%
  group_by(`NTD ID`, `Mode`) %>%
  summarize(Expenses = sum(Expenses), .groups = 'drop')

# Merge fare revenue and expenses data
FINANCIALS <- inner_join(FARES, EXPENSES, by = c("NTD ID", "Mode"))

# Load ridership data (UPT and VRM)
TRIPS <- read_xlsx("ridership.xlsx", sheet = "UPT") %>%
  filter(`Mode/Type of Service Status` == "Active") %>%
  select(-`Legacy NTD ID`, -`Reporter Type`, -`Mode/Type of Service Status`, -`UACE CD`, -`TOS`) %>%
  pivot_longer(-c(`NTD ID`, `Agency`, `UZA Name`, `Mode`, `3 Mode`), 
               names_to = "month", 
               values_to = "UPT") %>%
  drop_na() %>%
  mutate(month = lubridate::my(month))

MILES <- read_xlsx("ridership.xlsx", sheet = "VRM") %>%
  filter(`Mode/Type of Service Status` == "Active") %>%
  select(-`Legacy NTD ID`, -`Reporter Type`, -`Mode/Type of Service Status`, -`UACE CD`, -`TOS`) %>%
  pivot_longer(-c(`NTD ID`, `Agency`, `UZA Name`, `Mode`, `3 Mode`), 
               names_to = "month", 
               values_to = "VRM") %>%
  drop_na() %>%
  mutate(month = lubridate::my(month))

# Combine ridership data (UPT and VRM)
USAGE <- inner_join(TRIPS, MILES, by = c("NTD ID", "Agency", "UZA Name", "Mode", "month"))

# Ensure NTD ID is of the same type (character) in both data frames
USAGE <- USAGE %>%
  mutate(`NTD ID` = as.character(`NTD ID`))

FINANCIALS <- FINANCIALS %>%
  mutate(`NTD ID` = as.character(`NTD ID`))

# Combine usage data with financials
USAGE_AND_FINANCIALS <- left_join(USAGE, FINANCIALS, by = c("NTD ID", "Mode"))
```

Table Summarization

```{r}
# Filter the data for the year 2022 and summarize UPT and VRM
USAGE_2022_ANNUAL <- USAGE %>%
  filter(year(month) == 2022) %>%
  group_by(`NTD ID`, `Agency`, `UZA Name`, `Mode`) %>%
  summarize(
    Total_UPT = sum(UPT, na.rm = TRUE),
    Total_VRM = sum(VRM, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  rename(metro_area = `UZA Name`)  # Rename 'UZA Name' to 'metro_area'

# View the first few rows of the summarized data
head(USAGE_2022_ANNUAL)
```

Analysis

1. Which Transit System Had the Most UPT in 2022?
The transit system with the most Unlinked Passenger Trips (UPT) is:
```{r}
# Calculate the system with the most UPT in 2022
most_upt <- USAGE_2022_ANNUAL %>%
  group_by(`NTD ID`, `Agency`, `Mode`) %>%
  summarize(Total_UPT = sum(Total_UPT, na.rm = TRUE), .groups = 'drop') %>%
  arrange(desc(Total_UPT)) %>%
  slice(1)

most_upt
```

2. Which Transit System Had the Highest Farebox Recovery?
```{r}
# Calculate the system with the highest farebox recovery
highest_farebox <- USAGE_AND_FINANCIALS %>%
  mutate(Farebox_Recovery = `Total Fares` / Expenses) %>%
  arrange(desc(Farebox_Recovery)) %>%
  slice(1)

highest_farebox
```
3. Which Transit System Has the Lowest Expenses per UPT?
```{r}
# Calculate the system with the lowest expenses per UPT
lowest_exp_per_upt <- USAGE_AND_FINANCIALS %>%
  mutate(Expenses_per_UPT = Expenses / UPT) %>%
  arrange(Expenses_per_UPT) %>%
  slice(1)

lowest_exp_per_upt
```
4. Which Transit System Has the Highest Fares per UPT?
```{r}
# Calculate the system with the highest total fares per UPT
highest_fares_per_upt <- USAGE_AND_FINANCIALS %>%
  mutate(Fares_per_UPT = `Total Fares` / UPT) %>%
  arrange(desc(Fares_per_UPT)) %>%
  slice(1)

highest_fares_per_upt
```
5. Which Transit System Has the Lowest Expenses per VRM?
```{r}
# Calculate the system with the lowest expenses per VRM
lowest_exp_per_vrm <- USAGE_AND_FINANCIALS %>%
  mutate(Expenses_per_VRM = Expenses / VRM) %>%
  arrange(Expenses_per_VRM) %>%
  slice(1)

lowest_exp_per_vrm
```
6. Which Transit System Has the Highest Total Fares per VRM?
```{r}
# Calculate the system with the highest total fares per VRM
highest_fares_per_vrm <- USAGE_AND_FINANCIALS %>%
  mutate(Fares_per_VRM = `Total Fares` / VRM) %>%
  arrange(desc(Fares_per_VRM)) %>%
  slice(1)

highest_fares_per_vrm
```
Conclusion
Based on the analysis, I believe, the most efficient transit system varies depending on which factor of efficiency is prioritized:

Farebox recovery: Butler County Regional Transit Authority (Commuter Bus) demonstrates the highest farebox recovery, meaning it operates with a high degree of financial self-sustainability.
Operational Scale: MTA New York City Transit (Heavy Rail) handles the most UPT and VRM, making it highly efficient in serving a large population.








